---
title: "$\\LaTeX{}$ in R package documentation with mathjaxr"
author: "Mikael Jagan"
date: "07/06/2020"
output: rmarkdown::github_document
---

This document contains evolving notes on the
[mathjaxr](https://github.com/vnijs/MathJaxR)
package, developed by Wolfgang Viechtbauer
in response to mailing list threads 
[1](https://stat.ethz.ch/pipermail/r-devel/2020-April/079395.html)
and
[2](https://stat.ethz.ch/pipermail/r-package-devel/2020q2/005420.html).
The package automates integration of 
[MathJax](https://en.wikipedia.org/wiki/MathJax) 
in R package documentation,
making it much easier to get typeset math in HTML help pages.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Installation

The current development version of mathjaxr (1.0-0) is much more useful than 
current CRAN version (0.8-3). This will change with the next update, so check 
[here](https://cran.r-project.org/web/packages/mathjaxr/index.html) before 
installing.

```{r install}
# From CRAN
install.packages("mathjaxr")

# From GitHub (recommended until release of 1.0-0)
remotes::install_github("wviechtb/mathjaxr")
```

## Configuring `DESCRIPTION` and `NAMESPACE`

The mathjaxr package defines macros that you will use to type math into your
(roxygen2 or vanilla) documentation. To gain access to these macros, add the
following line to your `DESCRIPTION` file.

```{r description}
RdMacros: mathjaxr
```

To ensure that users who do not have mathjaxr installed get documentation with
correctly typeset math, add mathjaxr to `Suggests` or `Imports`. If you add it
to `Suggests`, and the user does not have mathjaxr installed, then MathJax will 
be sourced from a CDN (via the Internet) instead of locally from their mathjaxr 
installation. This means that math will **not** appear typeset if the user views
the HTML help while offline. To ensure that the user sees typeset math even 
while offline, you must add mathjaxr to `Imports`.

Since none of the functions in your R package will actually call functions from
mathjaxr, `R CMD check` and `devtools::check()` will produce the following note.

```{r check-note}
All declared Imports should be used
```

To suppress the note, add the following line to your `NAMESPACE` file.

```{r namespace}
import(mathjaxr)
```

## Configuring your documentation files

You must include `\loadmathjax` in all .R (roxygen2) or .Rd (vanilla) files
in which you'd like to type math. If using roxygen2, then configure your 
description section like so:

```{r loadmathjax-roxygen2}
#' Title text
#' 
#' \loadmathjax
#' Description text
```

(As usual, the `@description` tag has been omitted.) If editing an .Rd file 
directly, then this configuration should work:

```{r loadmathjax-vanilla}
\description{
  \loadmathjax
  Description text
}
```

Loading MathJax in the description section (as instructed by the maintainer)
does not prevent you from typing math in other sections, including `@title`, 
`@details`, and `@param`. However, `devtools::document()` will warn

```{r inline-warning}
@tag Use of inline HTML is not currently supported
```

if you have typed inline (not display) math outside of the description section.
I've been ignoring these warnings, because the math appears correctly typeset in 
the HTML help regardless of the section it is in.

Contradicting the maintainer's instructions, I have been using this (roxygen2) 
configuration without issue:

```{r loadmathjax-me}
#' \loadmathjax
#' Title text
#' 
#' Description text
```

Use at your own risk. I like it because it's preamble-y.

## Typing math

The workhorse macros are `\mjeqn{}{}` and `\mjdeqn{}{}` for inline math and 
display math, respectively. The first argument is the desired $\LaTeX{}$ code 
*without* dollar (\$) delimiters. The second argument is the desired plain text 
alternative, which will appear in the plain text (not HTML) help. If these two
arguments happen to be identical (perhaps you are just typing $a+b$, or perhaps 
you are fine with raw $\LaTeX{}$ code in the plain text help), then you can use 
the single argument macros `\mjseqn{}` and `\mjsdeqn{}` instead.

Here is a simple example from my package oneRous, which implements several 
methods for calculating or approximating unity.

```{r sumnegpowtwo}
#' Partial sum of negative powers of two
#' 
#' \loadmathjax
#' Calculates
#' \mjdeqn{\sum_{i=1}^{n} \frac{1}{2^i}}{1/2 + ... + 1/(2^n)}
#' for an integer \mjeqn{n \geq 1}{n >= 1}.
#' 
#' Returns unity (1) starting at \mjseqn{n = 54} due to convergence to within
#' machine precision. See [base::.Machine].
#' 
#' @param n An integer scalar. Must be positive.
#' @return A numeric scalar. The value of `sum(cumprod(rep(0.5, n)))`.
#' @export
#' @md
```

A list of supported $\LaTeX{}$ macros and environments can be found
[here](https://docs.mathjax.org/en/latest/input/tex/macros/index.html).

I have found that macro and environment usage within `\mjeqn{}{}` and friends 
can differ from typical usage. For example, I have used the `align` environment
successfully, but I needed to type newlines as `\cr` instead of `\\`. To get

$$\begin{align} x &= r \cos \theta \\ y &= r \sin \theta \end{align}$$
in the HTML help, I would write something like this:

```{r align}
#' \mjsdeqn{\begin{align} x &= r \cos\theta \cr y &= r \sin\theta \end{align}}
```

The need to type `\cr` instead of `\\` seems to apply more broadly (i.e., in 
other multiline environments like `bmatrix`).

## Previewing help without building

Given all of the issues with mathjaxr (see below), it is helpful to be able to 
view your package documentation to verify that your math was typeset, without 
needing to build and reinstall your package. The mathjaxr function 
`preview_rd()` allows you to preview documentation. For example, after editing 
`my_function.R`, you can view the updated HTML help in your browser by running

```{r preview-rd}
devtools::document()
mathjaxr::preview_rd(Rdfile = "my_function.Rd", type = "html")
```

You can view the updated plain text help with `type = "txt"`.

Confusingly, the first argument is just an .Rd file name, not a path to an .Rd
file. You must ensure that your working directory is `man/` or `man/..`, and 
specify `Rdfile = "my_function.Rd"` regardless.

## Known issues

Here is an evolving list of issues. Some are mentioned by the maintainer
[on GitHub](https://github.com/wviechtb/mathjaxr), while others I have
discovered. It's possible that my issues have workarounds that I just don't
know about.

### Strict inequalities

You must type strict inequalities with spaces (`i < j`, not `i<j`), or suffer
the wrath of the HTML interpreter.

### Literal braces and brackets

Literal braces must be typed as `\lbrace` and `\rbrace` instead of `\{` and 
`\}`. Literal brackets must be typed as `\lbrack` and `\rbrack` instead of `[` 
and `]`.

### Breaking long lines

Breaking long lines of $\LaTeX{}$ code over multiple lines in my source file
caused the typeset math to vanish. I have addressed this simply by ensuring that
any instance of a mathjaxr macro (e.g., `\mjsdeqn{}`) starts and ends on the 
same line.

```{r no-line-breaking}
#' Good
#' \mjsdeqn{\begin{align} x &= r \cos\theta \cr y &= r \sin\theta \end{align}}
#' 
#' Bad
#' \mjsdeqn{
#'   \begin{align} 
#'     x &= r \cos\theta \cr 
#'     y &= r \sin\theta 
#'   \end{align}
#' }
```

### No newlines in plain text

I have not figured out how to type newlines in plain text, making it difficult
to get something like a system of equations in the plain text help. Typical 
escapes like `\n` and `\cr` haven't worked for me.

```{r no-plain-newlines}
#' Good in HTML help, bad in plain text help
#' \mjdeqn{\begin{align} x - 2y &= 1 \cr 4x + 3y &= 5 \end{align}}{ x - 2y = 1 \cr 4x + 3y = 5}
```

There might be a (cumbersome) workaround involving the `\ifelse` macro, 
documented
[here](https://cran.r-project.org/doc/manuals/r-devel/R-exts.html#Conditional-text).

### Plain text isn't protected

When I enable Markdown with `@md`, my plain text gets processed like Markdown.
For example, I wanted `beta*S*I`, but got `beta\emph{S}I` in the plain text 
help (literally).

### Macros with subscripts, subscripts with macros (sometimes?)

Some of my equations were typeset fine until I tried to include something 
like $\widehat{N}_0$ (`\widehat{N}_0`), $\beta_\phi$ (`\beta_\phi`), or
$Z_\text{cum}$ (`Z_\text{cum}`), at which point raw $\LaTeX{}$ code appeared 
in the HTML help instead of typeset math. Curiously, those three terms 
are typeset fine individually. Including them as part of a longer line of 
$\LaTeX{}$ code is what seems to, in some instances but not others, break 
the interpreter.

I haven't deduced exactly what's going wrong, but I have found that enclosing
all of the $\LaTeX{}$ code in `\out{}` solves the issue, except for the fact
that "\\out" is also typeset (literally) before the desired math.

```{r subscripts}
#' Sometimes good, sometimes bad
#' \mjeqn{Z_\text{cum}}{Z_cum}
#'
#' Good, but "\out" is also typeset
#' \mjeqn{\out{Z_\text{cum}}}{Z_cum}
```

I suspect that there is a workaround without side effects.
